% !TEX root = 02.tex

\section{Documentation}
\begin{frame}[t]{LLVM official documentation}
  \begin{center}
    \begin{Huge}
      \vfill
      \url{llvm.org/docs}
      \vfill
    \end{Huge}
  \end{center}
\end{frame}
%--- Next Frame ---%

\begin{frame}[t]{A lot of documentation...}
  \url{llvm.org/docs} mentions:
  \begin{itemize}
    \item \texttt{\ 5} references about \textit{Design \& Overview}
    \item \texttt{23} references about \textit{User Guides}
    \item \texttt{15} references about \textit{Programming Documentation}
    \item \texttt{42} references about \textit{Subsystem Documentation}
    \item \texttt{\ 9} references about \textit{Development Process Documentation}
    \item \texttt{\ 5} Mailing Lists
    \item \texttt{\ 5} IRC bots
  \end{itemize}
  \vfill
  Most of the above references are OUT-OF-DATE.
  \vfill
  You probably need documentation about the documentation itself.
  \vfill
\end{frame}
%--- Next Frame ---%

\begin{frame}[t]{Essential documentation}
  \begin{description}
    \item[Intro to LLVM] \cite{LOCAL:www/llvmIntro}
          gives a quick and clear introduction to the compiler infrastructure.
          It is mostly up-to-date.\footnote{at the time I am writing}
    \vfill
    \item[Writing an LLVM pass] \cite{LOCAL:www/llvmWritingAPass}
          explains step by step how to implement a Pass
          for those who never did anything like that.
          We will see this tutorial later in the course.
    \vfill
    \item[Doxygen] \cite{LOCAL:www/llvmDoxygen}
          \textit{The best code documentation is the code itself.}
          Sometimes the generated doxygen documentation is enough.
          It also contains links to the web version of the source code.
          It is updated to the latest development branch.
          Please refer to github branches for the documentation about the stable versions.
    \vfill
    \item[llvm-dev] Mailing List. Last resource: ask other developers.
          Warning: 24/7 many people are posting in this ML.
  \end{description}
\end{frame}
